A Computational Account of Default-Mode Function\\by Control Theory and Reinforcement Learning

### 2016-08-15 paper/paper.tex
some new stuff

% macros from michael's .tex
\DeclareMathOperator{\dist}{dist} % The distance.
\DeclareMathOperator{\argmin}{argmin}
\DeclareMathOperator{\argmax}{argmax}
\DeclareMathOperator{\Id}{Id}
\DeclareMathOperator{\abs}{abs}
\newcommand{\R}{\mathbb{R}}
\newcommand{\N}{\mathbb{N}}
\newtheorem{thm}{Theorem}[section]
\newtheorem{prop}[thm]{Proposition}
\newtheorem{lem}[thm]{Lemma}
\newtheorem{cor}[thm]{Corollary}

@@

\subsubsection{Helmholtz free-energy and the generative model}
\begin{table}[H]
  \begin{tabular}{p{2cm}|p{11cm}}
         \hline
         \textbf{symbol}    & \textbf{description}  \\ \hline
         $\langle X\rangle_p$ & Expectation (a.k.a average, a.k.a mean value) of the
         random quantity $X$ w.r.t to the probability density $p$, formally defined by $\langle E\rangle_p := \sum_{z}p(z)X(z)$.\\ \hline
         $\mathcal H(p)$ & Information-theoretic entropy of a probability density $p$, formally defined by $\mathcal H(p) := -\sum_{z}p(z)\log(p(z))$,
          with the usual convention $0 \log(0) := 0$.\\ \hline
         $D_{KL}(q||p)$ & The Kullback-Leibler divergence between the probability densities $q$ and $p$ respectively, formally defined by $D_{KL}(q||p) := \sum_{z}q(z)\log(q(z)/p(z))$.\\ \hline
             $\zeta$ & Observations. In Friston's free-energy principle this has a decomposition in to two terms: the brain's internal state $b$ and sensory inputs $s$, i.e $\zeta = (s, b).$ \\ \hline
             $\psi$ & Hidden variables. This should be understood as the unobservable states of the external environment (to which the brain is trying to adapt by learning).\\ \hline
             $p_G(.|\zeta)$ & Generative density for ...\\ \hline
         $p_R(.|\zeta)$ & Recognition density for ... Does some kind of predictive coding (?).\\ \hline
         $F_G(\zeta)$ & Helmholtz free-energy for a model $p_G$ of generating the observation $\zeta$. This measures the surprise incured upon observing $\zeta$ generated by the model $G$.\\ \hline
         $F^R_G(\zeta)$ & Variational Helmholtz free-energy from model $R$
         to $G$.  Note that $F^G_G = F_G$.\\ \hline
  \end{tabular}
  \caption{Table of notations.}
\end{table}
Our starting point will be to build an approximation $p_G$ for the true density $p$ of the observations, so that this approximate density corresponds to the partition function of thermodynamic system. So,
\begin{eqnarray}
  \begin{split}
    \text{generative surprise } &= -\log(p_G(\zeta)) = -\log(p_G(\zeta)) \times 1 = -\log(P_G(\zeta))\sum_{\psi}p_G(\psi |\zeta)\\
    &= -\sum_{\psi}p_G(\psi, \zeta)\log(p_G(\zeta))
    =-\sum_{\psi}p_G(\psi |\zeta)\log(p_G(\psi, \zeta)/p_G(\psi|\zeta))\\
    &= \sum_{\psi}p_G(\psi |\zeta)\log(p_G(\psi|\zeta))-\sum_{\psi}p_G(\psi |\zeta)\log  (p_G(\psi, \zeta))\\
    &= -\langle \log  (p_G(., \zeta)) \rangle_{p_G(. |\zeta)} - \mathcal H(p_G(. |\zeta))\\
    &= \langle E_G(., \zeta) \rangle_{p_G(. |\zeta)} - \mathcal H(p_G(. |\zeta)) := F_G(\zeta)
  \end{split}
  \label{eq:helm}
\end{eqnarray}
where $E_G(\psi, \zeta)$ is the energy at \textit{macrostate} $\psi$ of a fictive thermodynamic system defined by setting
\begin{equation}
  E_G(\psi, \zeta) := -\log(p_G(\psi, \zeta)).
  \label{eq:gibbs}
\end{equation}
%% In the above, the variable $\psi$ is a hidden variable, and can / should be interpreted as the unobservable state of the external world.
The last quantity in \eqref{eq:helm} defined formally as
\begin{equation}
  F_G(\zeta) := \langle E_G(., \zeta) \rangle_{p_G(. |\zeta)} - \mathcal H(p_G(. |\zeta))
\end{equation}
is the \textit{Helmholtz free-energy}. %% , modelled as the physical system for which $p_G(\psi|\zeta)$ dictates the occupation probabilities of macrostates $\psi$
Thus generative surprise and generative Helmholtz free-energy are different views on exactly the same object.

@@

The following critics can be made:
\begin{itemize}
  \item 
    As noted by Dayan et al. (\textit{Variants of Helmholtz machines}), the inter-neuronal intra-layer independence assumption which is at the center of the HM becomes severely problematic as it is agnostic to the known organization of cortical layers...
    \item A drawback of the wake-sleep algorithm is that it requires a concurrent models (generative and recognitiion), which together do not correspond to optimization of (a bound of) the marginal likelihood (because of the incorrect KL used therein, etc.).
    \end{itemize}

\subsection{Free-energy minimization via bayesian auto-encoders}
Following Kingma et al. 2014 \cite{kingma2013auto}, define the data-dependent auxiliary random function
\begin{eqnarray}
  f_{G,R}(., \zeta) :\psi \mapsto \log(p_G(\psi,\zeta)) - \log(p_R(\psi|\zeta)).
\end{eqnarray}
Then we can rewrite the
variational free-energy as
\begin{eqnarray*}
  \begin{split}
    F_G^R(\zeta) &:= \langle E_G(., \zeta) \rangle_{p_R(. |\zeta)} - \mathcal H(p_R(. |\zeta)) = \langle E_G(., \zeta) + \log(p_R(.|\zeta))\rangle_{P_R(.|\zeta)}\\
    &=\langle -\log(p_G(.,\zeta)) + \log(p_R(.|\zeta))\rangle_{P_R(.|\zeta)}\\
    &= -\langle f_{G,R}\rangle_{P_R(.|\zeta)} \approx -\frac{1}{M}\sum_{m=1}^Mf_{G,R}(\psi^{(m)}), \text{ with }\psi^{(1)},\ldots,\psi^{(M)} \sim p_R(.|\zeta), \text{ and }M \rightarrow \infty.
    \end{split}
\end{eqnarray*}

\begin{mdframed}
  \textbf{Problem:} How do we sample from the recognition density $p_R(.|\zeta)$ in such a way that the sampling process is differentiable w.r.t the weights of the recognition network $\W^R$ ?
\end{mdframed}
\paragraph{Solution: The reparametrization trick.}
\begin{itemize}
\item Choose $\epsilon \sim p_{\text{noise}}$ (noise distribution, independent of $\W^R$!)
\item Set $\phi = g(\W^{R}, \zeta, \epsilon)$, where $g$ is an appropriate smooth function
  \begin{itemize}
  \item results in a sample $\phi \sim p_R(.|\zeta)$, from the correct
    posterior
    \end{itemize}
\end{itemize}

\paragraph{Examples.}
\begin{table}[H]
  \begin{tabular}{p{2cm}|p{1.5cm}|p{1.5cm}|p{2.2cm}|p{5cm}}
         \hline
         Example & $p_R(.|\zeta)$ & $p_{\text{noise}}$ & $g(\W^R,\zeta,\epsilon)$ & Also \\ \hline
         Normal dist. & $\mathcal N(\mu,\sigma)$ & $\mathcal N(0, 1)$ & $\mu + \sigma \odot \epsilon$ & Location-scale family: Laplace, Elliptical,
         Student’s t, Logistic, Uniform, Triangular, ... \\ \hline
         Exponential & $\mathcal \exp(\lambda)$ & $\mathcal U(0, 1)$ & $-\log(1-\epsilon)/\lambda$ & Invertible CDF: Cauchy, Logistic, Rayleigh, Pareta, Weibull, Reciprocal, Gompert, Gumbel, Erlan, ... \\ \hline
         Other & $\log\mathcal N(\mu,\sigma)$ & $\mathcal N(0, 1)$ & $\exp(\mu + \sigma \odot \epsilon)$ & Gamma, Dirichlet, Beta, Chi-squared, F, ... \\ \hline
  \end{tabular}
  \caption{Reparametrization trick for a variety of models. Kingma et al. 2014 \cite{kingma2013auto}.}
\end{table}

>>>>

% some useful macros
\DeclareMathOperator{\dist}{dist} % The distance.
\DeclareMathOperator{\argmin}{argmin}
\DeclareMathOperator{\argmax}{argmax}
\DeclareMathOperator{\Id}{Id}
\DeclareMathOperator{\abs}{abs}
\newcommand{\R}{\mathbb{R}}
\newcommand{\N}{\mathbb{N}}
\newtheorem{thm}{Theorem}[section]
\newtheorem{prop}[thm]{Proposition}
\newtheorem{lem}[thm]{Lemma}
\newtheorem{cor}[thm]{Corollary}

@@

\subsubsection{Helmholtz free-energy and the generative model}
\begin{table}[H]
  \begin{tabular}{p{2cm}|p{11cm}}
         \hline
         \textbf{symbol}    & \textbf{description}  \\ \hline
         $\langle X\rangle_p$ & Expectation (a.k.a average, a.k.a mean value) of the
         random quantity $X$ w.r.t to the probability density $p$, formally defined by $\langle E\rangle_p := \sum_{z}p(z)X(z)$.\\ \hline
         $\mathcal H(p)$ & Information-theoretic entropy of a probability density $p$, formally defined by $\mathcal H(p) := -\sum_{z}p(z)\log(p(z))$,
          with the usual convention $0 \log(0) := 0$.\\ \hline
         $D_{KL}(q||p)$ & The Kullback-Leibler divergence between the probability densities $q$ and $p$ respectively, formally defined by $D_{KL}(q||p) := \sum_{z}q(z)\log(q(z)/p(z))$.\\ \hline
             $\zeta$ & Observations. In Friston's free-energy principle this has a decomposition in to two terms: the brain's internal state $b$ and sensory inputs $s$, i.e $\zeta = (s, b).$ \\ \hline
             $\psi$ & Hidden variables. This should be understood as the unobservable states of the external environment (to which the brain is trying to adapt by learning).\\ \hline
             $p_G(.|\zeta)$ & Generative density for ...\\ \hline
         $p_R(.|\zeta)$ & Recognition density for ... Does some kind of predictive coding (?).\\ \hline
         $F_G(\zeta)$ & Helmholtz free-energy for a model $p_G$ of generating the observation $\zeta$. This measures the surprise incured upon observing $\zeta$ generated by the model $G$.\\ \hline
         $F^R_G(\zeta)$ & Variational Helmholtz free-energy from model $G$
          to $R$.  Note that $F^G_G = F_G$.\\ \hline
  \end{tabular}
  \caption{Table of notations.}
\end{table}
Our starting point will be to build an approximation $p_G$ for the true density $p$ of the observations, so that this approximate density corresponds to the partition function of thermodynamic system. So,
\begin{eqnarray}
  \begin{split}
    \text{generative surprise } &= -\log(p_G(\zeta)) = -\log(p_G(\zeta)) \times 1 = -\log(P_G(\zeta))\sum_{\psi}p_G(\psi |\zeta)\\
    &= -\sum_{\psi}p_G(\psi, \zeta)\log(p_G(\zeta))
    =-\sum_{\psi}p_G(\psi |\zeta)\log(p_G(\psi, \zeta)/p_G(\psi|\zeta))\\
    &= \sum_{\psi}p_G(\psi |\zeta)\log(p_G(\psi|\zeta))-\sum_{\psi}p_G(\psi |\zeta)\log  (p_G(\psi, \zeta))\\
    &= -\langle \log  (p_G(., \zeta)) \rangle_{p_G(. |\zeta)} - \mathcal H(p_G(. |\zeta))\\
    &= \langle E_G(., \zeta) \rangle_{p_G(. |\zeta)} - \mathcal H(p_G(. |\zeta)) := F_G(\zeta)
  \end{split}
  \label{eq:helm}
\end{eqnarray}
where $E_G(\psi, \zeta)$ is the energy at \textit{macrostate} $\psi$ of a fictive thermodynamic system defined by setting
\begin{equation}
  E_G(\psi, \zeta) := -\log(p_G(\psi, \zeta)).
  \label{eq:gibbs}
\end{equation}
%% In the above, the variable $\psi$ is a hidden variable, and can / should be interpreted as the unobservable state of the external world.
The last quantity in \eqref{eq:helm} defined formally as
\begin{equation}
  F_G(\zeta) := \langle E_G(., \zeta) \rangle_{p_G(. |\zeta)} - \mathcal H(p_G(. |\zeta))
\end{equation}
is the \textit{Helmholtz free-energy}. %% , modelled as the physical system for which $p_G(\psi|\zeta)$ dictates the occupation probabilities of macrostates $\psi$
Thus generative surprise and generative Helmholtz free-energy are different views on exactly the same object.

@@

The following critics can be made:
\begin{itemize}
  \item 
    As noted by Dayan et al. (\textit{Variants of Helmholtz machines}), the inter-neuronal intra-layer independence assumption which is at the center of the HM becomes severely problematic as it is agnostic to the known organization of cortical layers...
  \item A drawback of the wake-sleep algorithm is that it requires a concurrent models (generative and recognitiion), which together do not correspond to optimization of (a bound of) the marginal likelihood (because of the incorrect KL used therein, etc.).
  \item Also, note that the wake-sleep algorithm doesn't do backprop! This is due to
    echnical difficulty in getting derivatives of loss function w.r.t
    recognition wieghts $\W^R$).
  \item This difficulty was removed in the 2010s by Kingma et al., an other groups, via some kind of ``reparametrization trick''.

\subsection{Free-energy minimization via variational auto-encoders}
Following Kingma et al. 2014 \cite{kingma2013auto}, define the data-dependent auxiliary random function
\begin{eqnarray}
  f_{G,R}(., \zeta) :\psi \mapsto \log(p_G(\psi,\zeta)) - \log(p_R(\psi|\zeta)).
\end{eqnarray}
Then we can rewrite the
variational free-energy as
\begin{eqnarray*}
  \begin{split}
    F_G^R(\zeta) &:= \langle E_G(., \zeta) \rangle_{p_R(. |\zeta)} - \mathcal H(p_R(. |\zeta)) = \langle E_G(., \zeta) + \log(p_R(.|\zeta))\rangle_{P_R(.|\zeta)}\\
    &=\langle -\log(p_G(.,\zeta)) + \log(p_R(.|\zeta))\rangle_{P_R(.|\zeta)}\\
    &= -\langle f_{G,R}\rangle_{P_R(.|\zeta)} \approx -\frac{1}{M}\sum_{m=1}^Mf_{G,R}(\psi^{(m)}), \text{ with }\psi^{(1)},\ldots,\psi^{(M)} \sim p_R(.|\zeta), \text{ and }M \rightarrow \infty.
    \end{split}
\end{eqnarray*}

\begin{mdframed}
  \textbf{Problem:} How do we sample from the recognition density $p_R(.|\zeta)$ in such a way that the sampling process is differentiable w.r.t the weights of the recognition network $\W^R$ ?
\end{mdframed}
\paragraph{Solution: The reparametrization trick.}
\begin{itemize}
\item Choose $\epsilon \sim p_{\text{noise}}$ (noise distribution, independent of $\W^R$!)
\item Set $\phi = g(\W^{R}, \zeta, \epsilon)$, where $g$ is an appropriate class $\mathcal C^1$ function
  \begin{itemize}
  \item results in a sample $\phi \sim p_R(.|\zeta)$, from the correct
    posterior
    \end{itemize}
\end{itemize}

\begin{table}[H]
  \begin{tabular}{p{1.7cm}|p{1.5cm}|p{1.9cm}|p{2.1cm}|p{5cm}}
         \hline
         Posterior & $p_R(.|\zeta)$ & noise & $g(\W^R,\zeta,\epsilon)$ & Also \\ \hline
         Normal & $\mathcal N(\mu,\sigma)$ & $\epsilon \sim \mathcal N(0, 1)$ & $\mu + \sigma \odot \epsilon$ & Location-scale family: Laplace, Elliptical,
         Student’s t, Logistic, Uniform, Triangular, ... \\ \hline
         Exponential & $\mathcal \exp(\lambda)$ & $\epsilon \sim \mathcal U([0, 1))$ & $-\log(1-\epsilon)/\lambda$ & Invertible CDF: Cauchy, Logistic, Rayleigh, Pareta, Weibull, Reciprocal, Gompert, Gumbel, Erlan, ... \\ \hline
         Other & $\log\mathcal N(\mu,\sigma)$ & $\epsilon \sim \mathcal N(0, 1)$ & $\exp(\mu + \sigma \odot \epsilon)$ & Gamma, Dirichlet, Beta, Chi-squared, F, ... \\ \hline
  \end{tabular}
  \caption{Reparametrization trick for a variety of models. Kingma et al. 2014 \cite{kingma2013auto}.}
\end{table}

The result is a scheme for training DBNs via good-old backprop!
